Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-09-06 concurrency/CPP 2024/paper.tex
first pass at explaining maintenance ops

Unfortunately, while this specification suffices for the lock-coupling template, it cannot be used as is for the give-up template: because we allocate new empty leaf nodes, we must allocate \lstinline{node_t} structures, whose implementation in the give-up template has extra fields that do not appear in the lock-coupling template. As shown in Figure \ref{fig:insertOp_giveup}, a give-up template node (even an empty one!) has additional \lstinline{min} and \lstinline{max} fields, and these fields are needed to compute the bounds of the new child nodes. In a BST, if the current node's range is (\lstinline{min, max}) and its key is \lstinline{k}, its left child will have range (\texttt{min, k}) and its right child will have range (\texttt{k, max}). It may be possible to remedy this discrepancy by adding an ``allocate empty node'' function to the templates, which is then called by \lstinline{insertOp}, but this would still complicate the approach: originally, we expected that templates would be parameterized by data structure implementations but not vice versa. However, this two-way dependency seems fundamental when the layout of the node structure depends on the concurrency control mechanism we choose.

@@

\subsection{Templates vs. Internal Reorganization}
\label{internal_reorganize}
The \lstinline{delete} operation on binary search trees presents difficulties in applying the template approach. The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}). Likewise, in the data structures verified by Krishna et al., deletion can be performed by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation. When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
rotating three-node sections of the tree until the node to be deleted is at a leaf. This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).

It is instructive to compare BST deletion with Krishna et al.'s discussion of splitting in B-link trees. They assume the existence of a maintenance thread that constantly crawls the tree searching for full nodes and splitting them. The splitting can be defined as a sequential, data-structure-specific operation on a node and its parent; the crawl itself is a variant of the \lstinline{traverse} operation, and while it is not verified in their existing proofs, we expect that it could be defined as a template function. We might similarly imagine refactoring a lock-coupling \lstinline{pushdown_left} into a template-level traversal and a data-structure-level rotation operation. However, in the BST, rotation involves three nodes (the node to be deleted and both of its children)---and, crucially, \emph{we only know that it involves three nodes once we know the target data structure is a BST}. Thus, there is no clear way to define a template function that acquires enough locks (or otherwise performs enough synchronization) to call out to a sequential rotation operation without making the template specific to BSTs.

We believe this represents a fundamental limitation in the template approach. Operations in the style of \lstinline{pushdown_left} are ubiquitous in tree structures (e.g., the rebalancing operations of AVL and red-black trees), and always weave together synchronization and restructuring in a way that cannot easily be decomposed. This is a serious impediment to the goal of verifying concurrency control separately from the data structure: even if we have a verified template and a verified single-threaded tree, we still need to verify a complex data-structure-specific concurrent function to support deletion, which is often the most complicated part of a concurrent search structure.
%what's the conclusion here? can we see a way through?

@@

The total lines of code required were about the same as for the original VST verification of the BST. The vast majority of the proof effort was in the templates, in both \lstinline{traverse} and the top-level data structure operations. This is fairly encouraging, since this proof effort should be reusable for other data structures that use the same templates. For lock-coupling deletion, we retained the original (non-templated) proofs; we did not verify give-up deletion, since optimistic deletion is considerably harder to reason about and did not stand to benefit from the template approach.
%Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process. Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
%However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.  The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort. 
Proving the correctness of e.g. a lock-coupling linked list should be as simple as defining a new \lstinline{node} struct, and then implementing and verifying \lstinline{findNext} and \lstinline{insertOp} (plus the appropriate functions for deletion).
%Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct. While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.
The full development of our mechanization effort is available online at

\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.

>>>>

Unfortunately, while this specification suffices for the lock-coupling template, it cannot be used as is for the give-up template: because we allocate new empty leaf nodes, we must allocate \lstinline{node_t} structures, whose implementation in the give-up template has extra fields that do not appear in the lock-coupling template. As shown in Figure \ref{fig:insertOp_giveup}, a give-up template node (even an empty one!) has additional \lstinline{min} and \lstinline{max} fields, and these fields are needed to compute the bounds of the new child nodes. In a BST, if the current node's range is (\lstinline{min, max}) and its key is \lstinline{k}, its left child will have range (\texttt{min, k}) and its right child will have range (\texttt{k, max}). It may be possible to remedy this discrepancy by adding an ``allocate empty node'' function to the templates, which is then called by \lstinline{insertOp}, but this would still complicate the approach: originally, we expected that templates would be parameterized by data structure implementations but not vice versa. However, this two-way dependency seems fundamental when the layout of the node structure depends on the concurrency control mechanism we choose. For now, we break the abstraction boundary and verify two different versions of \lstinline{insertOp}, one for the lock-coupling template and one for the give-up template

@@

\subsection{Templates and Internal Reorganization}
\label{internal_reorganize}
The \lstinline{delete} operation on binary search trees highlights a non-obvious aspect of the template approach. The operations we have seen thus far can be factored into a data-structure-agnostic concurrent step (\lstinline{traverse}) and a concurrency-unaware data structure step (\lstinline{insert} or \lstinline{lookup}). In most of the data structures verified by Krishna et al., deletion can decomposed similarly, by traversing to the node to be deleted, acquiring its lock and possibly its parent's lock, and then performing a local, sequential data-structure-specific operation. When removing a node from a BST, on the other hand, we usually restructure the tree with a \lstinline{pushdown_left} operation, %include figure or code?
rotating three-node sections of the tree until the node to be deleted is at a leaf. This operation is simultaneously concurrency-aware (we must lock the nodes involved to avoid race conditions) and data-structure-specific (we need to know precisely which nodes to target and how to rearrange them).

Krishna et al. also discuss an operation of this sort, the \lstinline{split} operation in B-link trees. Their approach is to assume the existence of a separate maintenance thread that constantly crawls the tree searching for full nodes and splitting them. The split itself is a no-op on the abstract state of the data structure: it makes internal structural changes but does not change the set of keys stored in the tree. This complicates the description of the template approach, but such a maintenance operation is necessary for many concurrent data structures. In general, we believe that the template approach should be thought of as the combination of \emph{three} kinds of operations: concurrent, data-structure-agnostic template operations; sequential data structure operations; and no-op maintenance operations, which are both concurrency-aware and data-structure-specific, but whose specifications leave the abstract state unchanged. For instance, rotation in the BST can be seen as a maintenance operation, and the deletion of a node that has already been rotated to a leaf position is a sequential, data-structure-specific operation. Rebalancing operations in AVL and red-black trees would also be considered maintenance operations. This can be seen as a limitation of the template approach, since concurrency cannot be completely decoupled from data-structure-specific reasoning, but because maintenance operations are logical no-ops, they should be easier to verify than concurrent data structure operations in general. In future work, we hope to provide a formal account of these three categories of operations, and a theorem describing how the proofs of each kind of operation combine to yield a top-level verified data structure. This will be essential for applying the template approach to the complexity of real-world high-performance data structure implementations.

@@

The total lines of code required were about the same as for the original VST verification of the BST. The vast majority of the proof effort was in the templates, in both \lstinline{traverse} and the top-level data structure operations. This is fairly encouraging, since this proof effort should be reusable for other data structures that use the same templates. For lock-coupling deletion, we retained the original (non-templated) proofs; we did not verify give-up deletion, since optimistic deletion is considerably harder to reason about and did not stand to benefit from the template approach.
%Even though the proof for the \texttt{insert} function within the lock-coupling templates was line-of-code (LoC) intensive, future enhancements could significantly streamline this process. Otherwise, the construction of proofs for the \texttt{traverse} function, in the contexts of both lock-coupling and give-up templates, necessitated a considerable amount of time and effort.
%However, we expect that functions such as \texttt{traverse}, \texttt{insert}, and \texttt{lookup} will be reusable with only minor adjustments needed.  The methods \texttt{findNext} and \texttt{insertOp} may also be adapted to other data structures, such as linked lists, with some additional effort. 
Proving the correctness of e.g. a lock-coupling linked list should be as simple as defining a new \lstinline{node} struct, and then implementing and verifying \lstinline{findNext} and \lstinline{insertOp} (plus the appropriate functions for deletion).
%Specifically, when attempting to prove concurrent linked lists, we plan to tweak the specifications of \texttt{findNext}, \texttt{inRange}, and \texttt{insertOp} to restructure the \texttt{node} struct. While the size of our template proofs currently approximates that of the no-template versions, we anticipate that strategic application of simple tactics for quantifier expansion could significantly streamline these proofs in the future.
The full development of our mechanization effort is available online at [link redacted for anonymity].
%\href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}.
